Collaborative Multioutput Gaussian Processes: \\ Supplementary Material

### 2014-03-19 paper/supplemental.tex
update supplemental

\section{Gaussian Identities}
Let $\y$, $\g = \{\g_j\}_{j=1}^q$, and $\h$ be random variables with multivariate Gaussian distributions: 
$p(\y | \g, \h) = \Normal(\y; \sum_{j=1}^Q \W_j \g_j + \W \h, \beta^{-1} \I)$, $p(\g_j) = \Normal(\g_j; \m_j, \S_j)$, and $p(\h) = \Normal(\h; \m, \S)$.
The following identity is important in deriving the evidence lower bound:
\begin{align}
\nonumber
\int& \log p(\y | \g, \h) \prod_{j=1}^q p(\g_j) p(\h) \der \g \der \h \\
\nonumber
=& \log \Normal(\y; \sum_{j=1}^q \W_j \m_j + \W \m, \beta^{-1} \I) \\
& - \frac{1}{2} \beta \trace \W^T \W \S - \frac{1}{2} \beta \trace \sum_{j=1}^q \W_j^T \W_j \S_j.
 \label{eq:identity}
\end{align}
The identity can be proved by using this fact: 
\begin{align}
\nonumber
&\int (\W\x - \Mu)^T \BigSigma^{-1} (\W \x - \Mu) \Normal(\x; \m, \S) \der \x \\
&= (\Mu - \W\m)^T \BigSigma^{-1} (\Mu - \W\m) + \trace \W^T \BigSigma^{-1} \W \S.
\end{align}

\section{Expectation of an Individual Likelihood Term}
Applying the identity in eq. \ref{eq:identity}, the expectation of an individual likelihood term with respect to the posterior distribution is given by:
\begin{align}
\nonumber
l_{in} =& \int \log p(y_{in} | \g_n, h_{in}) \prod_{j=1}^Q p(g_{jn} | \u_j) p(h_{in} | \u_i) \der \g_n \der h_{in} \\
\nonumber
=& \log \Normal(y_{in}; \sum_{j=1}^Q w_{ij} \mu_{jn}+ \mu^h_{in}, \beta_i^{-1}) \\
&- \frac{1}{2} \beta_i \sum_{j=1}^Q w_{ij}^2 \tilde{k}_{nn} 
- \frac{1}{2} \beta_i \tilde{k}^h_{inn},
\label{eq:lin}
\end{align}
where $\tilde{k}_{nn} = (\tilde{\K})_{nn}$, $\tilde{k}^h_{inn} = (\tilde{\K}^h_i)_{nn}$, $\mu_{jn} = (\Mu_j)_n$, and $\mu^h_{in} = (\Mu^h_i)_n$.

@@

\section{Derivatives of the ELBO}
For convenience, we derive the gradients of the lower bound for the case of a single GP (i.e. the bound in \citet{hensmangaussian}).
The derivatives of the ELBO of the collaborative multioutput GPs model are  typically linear combination of the derivatives here.
The lower bound as a function of all parameters is
\begin{align}
\nonumber
\calL
=& \log \Normal(\y; \K_{NM} \K_{MM}^{-1} \m, \beta^{-1}\I) \\
\nonumber
 &- \frac{1}{2} \beta \trace \tilde{\K}
 - \frac{1}{2} \beta \trace (\S\K_{MM}^{-1} \K_{MN} \K_{NM} \K_{MM}^{-1}) \\  \nonumber
&- \frac{1}{2} \left( \log |\K_{MM}| + \trace(\K_{MM}^{-1}(\m \m^T + \S)) \right) \\ \nonumber
=& \underbrace{\log \Normal(\y; \A \m, \beta^{-1}\I)}_{\calL_1}
 - \underbrace{\frac{1}{2} \beta \trace (\K_{NN} - \A\K_{MN})}_{\calL_2} \\
 \nonumber
&- \underbrace{\frac{1}{2} \left( \log |\K_{MM}| + \trace(\K_{MM}^{-1}(\m \m^T + \S)) \right)}_{\calL_4}\\
 & - \underbrace{\frac{1}{2} \beta \trace (\S\A^T\A)}_{\calL_3},
\end{align}
where $\A = \K_{NM} \K_{MM}^{-1}$.
Here $\K_{NM} = k(\X,\Z)$ is the cross-covariance matrix between the observed inputs and the inducing inputs and $\K_{MM} = k(\Z,\Z)$ is the auto-covariance matrix between the inducing inputs.
Notice that we have re-written the sum of individual terms in matrix form which will make the derivation easier and also the computation faster via vectorization.

\subsection{Derivative of the Noise Hyperparameter}
The derivative of the noise hyperparameter $\beta$ is easily computed as:
\begin{align}
\deriv{\calL}{\beta} = \frac{N}{2\beta} - \frac{1}{2} (\y - \A\m)^T (\y - \A\m) - \frac{\calL_2}{\beta} - \frac{\calL_3}{\beta}.
\end{align}
\subsection{Derivatives of the Covariance Hyperparameters}  
To simplify the math, we utilize the matrix $\A$ defined above.
Firstly, the derivative of $\A$ wrt a covariance hyperparameter $t$ is given by:
\begin{align}
\deriv{\A}{t} = \left(\deriv{\K_{NM}}{t} - \A \deriv{\K_{MM}}{t}\right)\K_{MM}^{-1}.
\end{align}
The derivatives of $\calL_1, \calL_2, \calL_3 \text{ and } \calL_4$ are thus given by:
\begin{align}
\deriv{\calL_1}{t} =& \beta (\y - \A\m)^T \deriv{\A}{t} \m \\
\deriv{\calL_2}{t} =& \frac{1}{2}\beta \trace \left(\deriv{\K_{NN}}{t} - \A \deriv{\K_{MN}}{t} - \deriv{\A}{t} \K_{MN}\right) \\
\deriv{\calL_3}{t} =& \beta \trace \left(\A \S \deriv{\A^T}{t} \right) \\
\nonumber
\deriv{\calL_4}{t} =& \frac{1}{2}  \trace \left(\K_{MM}^{-1} \deriv{ \K_{MM}}{t}\right) \\
&- \frac{1}{2} \trace \left(\K_{MM}^{-1} \deriv{\K_{MM}}{t} \K_{MM}^{-1} (\m \m^T + \S) \right) 
\end{align}
The derivatives are then computed by taking the derivatives of the covariance matrices $\K_{NN} (\text{the diagonal only}), \K_{NM} \text{ and }  \K_{MM}$, hence the covariance function, wrt the hyperparameters.

@@

\noindent We re-write $\deriv{\calL_1}{t}, \deriv{\calL_2}{t}, \deriv{ \calL_3}{t}, \deriv{\calL_4}{t}$ by expanding $\deriv{\A}{t}$ (here $t = z_{mj}$):
% dL1
\begin{align}
\nonumber
\deriv{\calL_1}{t}
=& \beta \trace (\y - \A\m)^T \left(\deriv{\K_{NM}}{t} -  \A \deriv{ \K_{MM}}{t}\right)\K_{MM}^{-1} \m \\
\nonumber
=& \beta \trace \K_{MM}^{-1} \m (\y - \A\m)^T \deriv{\K_{NM}}{t} \\
&-\beta \trace \K_{MM}^{-1} \m (\y - \A\m)^T \A \deriv{\K_{MM}}{t}
\end{align}
%dL2
\begin{align}
\deriv{\calL_2}{t}
&= - \beta \trace \A^T \deriv{\K_{NM}}{t}
 + \frac{1}{2} \beta \trace \A^T \A \deriv{\K_{MM}}{t}
 \end{align}
% dL3
 \begin{align}
\deriv{\calL_3}{t}
&= \beta \trace \K_{MM}^{-1} \S \A^T \deriv{\K_{NM}}{t}
 - \beta \trace \K_{MM}^{-1} \S \A^T \A \deriv{\K_{MM}}{t}
 \end{align}
 % dL4
 \begin{align}
\deriv{\calL_4}{t}
 =& \frac{1}{2}  \trace \K_{MM}^{-1} \deriv{\K_{MM}}{t} \\
 \nonumber
& - \frac{1}{2} \trace \K_{MM}^{-1} (\m \m^T + \S) \K_{MM}^{-1} \deriv{\K_{MM}}{t} 
\end{align}
From the above 4 equations we get,
\begin{align}
\deriv{\calL}{t} = \trace \D_1 \deriv{\K_{NM}}{t} + \trace \D_2 \deriv{ \K_{MM}}{t},
\end{align}
where 
\begin{align}
\D_1 =& \beta \K_{MM}^{-1} \m (\y - \A\m)^T
 + \beta \A^T
 - \beta \K_{MM}^{-1} \S \A^T \\ \nonumber
\D_2 =& -\beta \trace \K_{MM}^{-1} \m (\y - \A\m)^T \A
 - \frac{1}{2} \beta \A^T \A -\frac{1}{2} \K_{MM}^{-1} \\
  &+ \beta \K_{MM}^{-1} \S \A^T \A	 
   + \frac{1}{2} \K_{MM}^{-1} (\m \m^T + \S) \K_{MM}^{-1}
\end{align}
%TODO: give the correct cost
Notice that $\D_1$ and $\D_2$ can be pre-computed with a cost of $\calO(M^3)$ (or $\calO(N_bM^2)$ if the minibatch size $N_b > M$).
The computational cost of taking derivatives of $MD$ inducing parameters is thus $\calO(M^3 + MDM) = \calO(M^3)$ as the cost of the two $\trace$ operators is $\calO(M)$ due to the fact that only $\calO(M)$ elements of $\deriv{\K_{MM}}{t}$ or $\deriv{\K_{NM}}{t}$ are non-zero.
This fact can be further exploited to perform vectorized operation, for e.g. in Matlab, such that the iteration over all inducing inputs can be avoided.

>>>>

\section{Gaussian Identities}
Let $\g = \{\g_j\}_{j=1}^q$, $\h$, and $\y$ be random variables with multivariate Gaussian distributions: 
$p(\y | \g, \h) = \Normal(\y; \sum_{j=1}^Q \W_j \g_j + \W \h, \beta^{-1} \I)$, $p(\g_j) = \Normal(\g_j; \m_j, \S_j)$, and $p(\h) = \Normal(\h; \m, \S)$.
The following identity is important in deriving the evidence lower bound:
\begin{align}
\nonumber
\int& \log p(\y | \g, \h) \prod_{j=1}^q p(\g_j) p(\h) \der \g \der \h \\
\nonumber
=& \log \Normal(\y; \sum_{j=1}^q \W_j \m_j + \W \m, \beta^{-1} \I) \\
& - \frac{1}{2} \beta \trace \W^T \W \S - \frac{1}{2} \beta \trace \sum_{j=1}^q \W_j^T \W_j \S_j.
 \label{eq:identity}
\end{align}
The identity can be proved by using this fact: 
\begin{align}
\nonumber
&\int (\W\x - \Mu)^T \BigSigma^{-1} (\W \x - \Mu) \Normal(\x; \m, \S) \der \x \\
&= (\Mu - \W\m)^T \BigSigma^{-1} (\Mu - \W\m) + \trace \W^T \BigSigma^{-1} \W \S.
\end{align}

\section{Derivation of the Variational Lower Bound}
The variational lower bound of the log marginal (eq. 13 in the main text) is given by:
\begin{align}
\nonumber
\log p(\y) \ge& \int q(\u, \v) \log \frac{p(\y | \u, \v) p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
\nonumber
%=& \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v  \\ 
%\nonumber
&+ \int q(\u, \v) \log \frac{p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
\nonumber
=& \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v \\
\nonumber
&- \sum_{j=1}^Q \KL[q(\u_j) || p(\u_j)] - \sum_{i=1}^P \KL[q(\v_i) || p(\v_i)],
\end{align}

Since the KL terms are analytically tractable, we compute the first term in the above sum. 
This is done first by deriving a lower bound to the likelihood $p(\y | \u, \v)$ (eq. 14 in the main text).
\begin{align}
\nonumber
\log \text{ } p(\y | \u, \v)
&= \log \Eb{p(\y | \g, \h)}_{p(\g,\h | \u, \v)} \\
\nonumber
&\ge \Eb{\log p(\y | \g, \h)}_{p(\g,\h | \u, \v)}  \\
\nonumber
&= \sum_{i=1}^P \sum_{n=1}^N \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g | \u) p(\h_i | \v_i)} 
\end{align}
Applying the identity in eq. \ref{eq:identity}, 
the expectation of an individual likelihood term with respect to the posterior distribution is given by: 
\begin{align}
\nonumber
l_{in} =& \int \log p(y_{in} | \g_n, h_{in}) \prod_{j=1}^Q p(g_{jn} | \u_j) p(h_{in} | \u_i) \der \g_n \der h_{in} \\
\nonumber
=& \log \Normal(y_{in}; \sum_{j=1}^Q w_{ij} \mu_{jn}+ \mu^h_{in}, \beta_i^{-1}) \\
&- \frac{1}{2} \beta_i \sum_{j=1}^Q w_{ij}^2 \tilde{k}_{jnn} 
- \frac{1}{2} \beta_i \tilde{k}^h_{inn},
\end{align}
where $\tilde{k}_{jnn} = (\tilde{\K}_j)_{nn}$, $\tilde{k}^h_{inn} = (\tilde{\K}^h_i)_{nn}$, $\mu_{jn} = (\Mu_j)_n$, and $\mu^h_{in} = (\Mu^h_i)_n$.

Substituting $l_{in}$ into the expression for the lower bound of $\log p(\y | \u, \v)$ (again, this is eq. 14 in the main text), and applying the identity in eq. \ref{eq:identity} to carry out the integral we obtain the lower bound as given in the main text.

@@

\section{Derivatives of the ELBO}
For exposition, we derive the gradients of the lower bound for the case of a single GP (i.e. the bound in \citet{hensmangaussian}).
The derivatives of the ELBO of the collaborative multioutput GPs model are typically linear combination of the derivatives here.
The lower bound as a function of all parameters is
\begin{align}
\nonumber
\calL
=& \log \Normal(\y; \K_{NM} \K_{MM}^{-1} \m, \beta^{-1}\I) \\
\nonumber
 &- \frac{1}{2} \beta \trace \tilde{\K}
 - \frac{1}{2} \beta \trace (\S\K_{MM}^{-1} \K_{MN} \K_{NM} \K_{MM}^{-1}) \\  \nonumber
&- \frac{1}{2} \left( \log |\K_{MM}| + \trace(\K_{MM}^{-1}(\m \m^T + \S)) \right) \\ \nonumber
=& \underbrace{\log \Normal(\y; \A \m, \beta^{-1}\I)}_{\calL_1}
 - \underbrace{\frac{1}{2} \beta \trace (\K_{NN} - \A\K_{MN})}_{\calL_2} \\
 \nonumber
&- \underbrace{\frac{1}{2} \left( \log |\K_{MM}| + \trace(\K_{MM}^{-1}(\m \m^T + \S)) \right)}_{\calL_4}\\
 & - \underbrace{\frac{1}{2} \beta \trace (\S\A^T\A)}_{\calL_3},
\end{align}
where $\A = \K_{NM} \K_{MM}^{-1}$.
Here $\K_{NM} = k(\X,\Z)$ is the cross-covariance matrix between the observed inputs and the inducing inputs and $\K_{MM} = k(\Z,\Z)$ is the auto-covariance matrix between the inducing inputs.
Notice that we have re-written the sum of individual terms in matrix form which will make the derivation easier and also the computation faster via vectorization.

\subsection{Derivative of the Noise Hyperparameter}
The derivative of the noise hyperparameter $\beta$ is easily computed as:
\begin{align}
\deriv{\calL}{\beta} = \frac{N}{2\beta} - \frac{1}{2} (\y - \A\m)^T (\y - \A\m) - \frac{\calL_2}{\beta} - \frac{\calL_3}{\beta}.
\end{align}
\subsection{Derivatives of the Covariance Hyperparameters}  
To simplify the math, we utilize the matrix $\A$ defined above.
Firstly, the derivative of $\A$ wrt a covariance hyperparameter $t$ is given by:
\begin{align}
\deriv{\A}{t} = \left(\deriv{\K_{NM}}{t} - \A \deriv{\K_{MM}}{t}\right)\K_{MM}^{-1}.
\end{align}
The derivatives of $\calL_1, \calL_2, \calL_3 \text{ and } \calL_4$ are thus given by:
\begin{align}
\deriv{\calL_1}{t} =& \beta (\y - \A\m)^T \deriv{\A}{t} \m \\
\deriv{\calL_2}{t} =& \frac{1}{2}\beta \trace \left(\deriv{\K_{NN}}{t} - \A \deriv{\K_{MN}}{t} - \deriv{\A}{t} \K_{MN}\right)
\end{align}
\begin{align}
\deriv{\calL_3}{t} =& \beta \trace \left(\A \S \deriv{\A^T}{t} \right) \\
\nonumber
\deriv{\calL_4}{t} =& \frac{1}{2}  \trace \left(\K_{MM}^{-1} \deriv{ \K_{MM}}{t}\right) \\
&- \frac{1}{2} \trace \left(\K_{MM}^{-1} \deriv{\K_{MM}}{t} \K_{MM}^{-1} (\m \m^T + \S) \right) 
\end{align}
The derivatives are then computed by taking the derivatives of the covariance matrices $\K_{NN} (\text{the diagonal only}), \K_{NM} \text{ and }  \K_{MM}$, hence the covariance function, wrt the hyperparameters.

@@

\noindent We re-write $\deriv{\calL_1}{t}, \deriv{\calL_2}{t}, \deriv{ \calL_3}{t}, \deriv{\calL_4}{t}$ by expanding $\deriv{\A}{t}$ (here $t = z_{mj}$):
% dL1
\begin{align}
\nonumber
\deriv{\calL_1}{t}
=& \beta \trace (\y - \A\m)^T \left(\deriv{\K_{NM}}{t} -  \A \deriv{ \K_{MM}}{t}\right)\K_{MM}^{-1} \m \\
\nonumber
=& \beta \trace \K_{MM}^{-1} \m (\y - \A\m)^T \deriv{\K_{NM}}{t} \\
&-\beta \trace \K_{MM}^{-1} \m (\y - \A\m)^T \A \deriv{\K_{MM}}{t}
\end{align}
%dL2
\begin{align}
\deriv{\calL_2}{t}
&= - \beta \trace \A^T \deriv{\K_{NM}}{t}
 + \frac{1}{2} \beta \trace \A^T \A \deriv{\K_{MM}}{t}
 \end{align}
% dL3
 \begin{align}
\deriv{\calL_3}{t}
&= \beta \trace \K_{MM}^{-1} \S \A^T \deriv{\K_{NM}}{t}
 - \beta \trace \K_{MM}^{-1} \S \A^T \A \deriv{\K_{MM}}{t}
 \end{align}
 % dL4
 \begin{align}
\deriv{\calL_4}{t}
 =& \frac{1}{2}  \trace \K_{MM}^{-1} \deriv{\K_{MM}}{t} \\
 \nonumber
& - \frac{1}{2} \trace \K_{MM}^{-1} (\m \m^T + \S) \K_{MM}^{-1} \deriv{\K_{MM}}{t} 
\end{align}
From the above 4 equations we get,
\begin{align}
\deriv{\calL}{t} = \trace \D_1 \deriv{\K_{NM}}{t} + \trace \D_2 \deriv{ \K_{MM}}{t},
\end{align}
where 
\begin{align}
\D_1 =& \beta \K_{MM}^{-1} \m (\y - \A\m)^T
 + \beta \A^T
 - \beta \K_{MM}^{-1} \S \A^T \\ \nonumber
\D_2 =& -\beta \trace \K_{MM}^{-1} \m (\y - \A\m)^T \A
 - \frac{1}{2} \beta \A^T \A -\frac{1}{2} \K_{MM}^{-1} \\
  &+ \beta \K_{MM}^{-1} \S \A^T \A	 
   + \frac{1}{2} \K_{MM}^{-1} (\m \m^T + \S) \K_{MM}^{-1}
\end{align}
%TODO: give the correct cost
Notice that $\D_1$ and $\D_2$ can be pre-computed with a cost of $\calO(M^3)$ (or $\calO(N_bM^2)$ if the minibatch size $N_b > M$).
The computational cost of taking derivatives of $MD$ inducing parameters is thus $\calO(M^3 + MDM) = \calO(M^3)$ as the cost of the two trace operators is $\calO(M)$ due to the fact that only $\calO(M)$ elements of $\deriv{\K_{MM}}{t}$ or $\deriv{\K_{NM}}{t}$ are non-zero.

For implementation with MATLAB, a loop over $M \times D$ parameters of the inducing inputs can be very slow for even moderate values of $M$ and $D$.
The aforementioned fact about $\deriv{\K_{MM}}{t}$ and $\deriv{\K_{NM}}{t}$ can be used to perform vectorized operations that compute the derivatives of all $M$ parameters given a specific dimension at a cost of $\calO(M^2)$.
The loop is the executed over the input dimension $D$, leading to a complexity of still $\calO(DM^2)$ only.
